Numerical simulations of the Cordilleran ice sheet through the last glacial cycle

### 2015-05-27 cordillera-cycle.tex
Update text and improve phrasing (Arjen).

More than a century of exploration and geological investigation of the
Cordilleran mountains have led to many observations in support of the former
ice sheet
    \citep{Jackson.Clague.1991}.
Despite the lack of documented end moraines offshore, in the zone of confluence
with the Laurentide ice sheet, and in areas swept by the Missoula floods
    \citep{Carrara.etal.1996},
moraines that demarcate the northern and south-western margins provide key
constraints that allow reasonable reconstructions of maximum ice sheet extents
    (\citealp{Prest.etal.1968}; \citealp[Fig. 1.12]{Clague.1989};
     \citealp{Duk-Rodkin.1999};
     \citealp{Booth.etal.2003}; \citealp{Dyke.2004}).
The LGM Cordilleran ice sheet extent was short-lived, as indicated by
field evidence from radiocarbon dating
    \citep{Clague.etal.1980, Clague.1985, Clague.1986, Porter.Swanson.1998,
           Menounos.etal.2008},
cosmogenic exposure dating
    \citep{Stroeven.etal.2010, Stroeven.etal.2014, Margold.etal.2014},
bedrock deformation in response to former ice loads
    \citep{Clague.James.2002, Clague.etal.2005},
and offshore sedimentary records
    \citep{Cosma.etal.2008, Davies.etal.2011}.
However, former ice thicknesses and, therefore, the ice sheet's contribution to
the LGM sea level low stand
    \citep{Carlson.Clark.2012, Clark.Mix.2002}
remain uncertain.

@@

\begin{figure}
  \includegraphics{locmap}
  \caption{Relief map of northern North America showing a reconstruction of the
           areas once covered by the Cordilleran (CIS), Laurentide (LIS),
           Innuitian (IIS) and Greenland (GIS) ice sheets during the last
           18\,\unit{\chem{^{14}C}\,ka} \citep[21.4\,cal\,ka,][]{Dyke.2004}.
           The rectangular box denotes the location of the
           modelling domain used in this study. Major mountain ranges covered
           by the ice sheet include the Alaska Range (AR), the Wrangell and
           St.-Elias mountains (WSEM), the Selwyn and Mackenzie mountains
           (SMKM), the Skeena Mountains (SM), the Coast and Rocky
           Mountains (CRM) and the North Cascades (NC). Major depressions
           include the Liard Lowland (LL), the Interior Plateau of British
           Columbia (IP) and the Puget Lowland (PL). The background
           map consists of ETOPO1 \citep{Amante.Eakins.2009} and Natural Earth
           Data \citep{Patterson.Kelso.2014}.}
  \label{fig:locmap}
\end{figure}

In contrast, evidence for the deglaciation history of the Cordilleran
ice sheet since the LGM is considerable, albeit mostly at a regional scale.
Geomorphological evidence from south-central British Columbia indicates a rapid
deglaciation, including an early emergence of the elevated areas while thin,
stagnant ice still covered the surrounding lowlands
    \citep{Fulton.1967, Fulton.1991, Margold.etal.2011, Margold.etal.2013a}.
This model, although credible, may not apply in all areas of the Cordilleran
ice sheet
    \citep{Margold.etal.2013}.
Although solid evidence for late-glacial glacier re-advances have been found in
the Coast, Columbia and Rocky mountains
    \citep{Clague.etal.1997, Friele.Clague.2002, Friele.Clague.2002a,
           Kovanen.2002, Kovanen.Easterbrook.2002, Lakeman.etal.2008,
           Menounos.etal.2008},
it appears to be more sparse than for formerly glaciated regions surrounding
the North Atlantic
    \citep[e.g.,][]{Sissons.1979, Lundqvist.1987,
                    Ivy-Ochs.etal.1999, Stea.etal.2011}.
Nevertheless, recent oxygen isotope measurements from Gulf of Alaska sediments
reveals a climatic evolution highly correlated to that of Greenland during this
period, including a distinct late glacial cold reversal between 14.1 and
11.7\,ka \citep{Praetorius.Mix.2014}.

In general, the topographic complexity of the North American Cordillera and its
effect on glacial history have inhibited the reconstruction of ice sheet-wide
glacial advance and retreat patterns such as
those available for the Fennoscandian and Laurentide ice sheets
     \citep{Boulton.etal.2001, Dyke.Prest.1987, Dyke.etal.2003,
            Kleman.etal.1997, Kleman.etal.2010}.
Here, we use a numerical ice sheet model \citep{PISM-authors.2014},
calibrated against field-based evidence, to perform a quantitative
reconstruction of the Cordilleran ice sheet history through the last glacial
cycle, and
analyse some of the long-standing questions related to its evolution:

@@

Although numerical ice sheet modelling has been established as a useful tool to
improve our understanding of the Cordilleran ice sheet
    (\citealp[p.~227]{Jackson.Clague.1991}; \citealp{Robert.1991},
     \citealp{Marshall.etal.2000}),
the ubiquitously mountainous
topography of the region has presented two major challenges to its application.
First, only recent development of numerical ice sheet models and underlying
scientific computing tools \citep{Bueler.Brown.2009, Balay.etal.2014} has
allowed for high-resolution numerical modelling of glaciers and ice sheets on
mountainous terrain
over millenial time scales \citep[e.g.,][]{Golledge.etal.2012}. Second, the
complex
topography of the North American Cordillera also induces strong gradients in
temperature
seasonality and distinct patterns of precipition, thus requiring the use of
high-resolution climate forcing fields as an input to the ice sheet model
\citep{Seguinot.etal.2014}.

@@

The simulations presented here were run using the Parallel Ice Sheet Model
(PISM, development version~8ff7cbe), an open source,
finite difference, shallow ice sheet model \citep{PISM-authors.2014}. The model
requires input on basal topography, sea level, geothermal heat flux and
climate forcing. It computes the evolution of ice extent
and thickness over time, the thermal and dynamic
states of the ice sheet, and the associated lithospheric response.

@@

A~pseudo-plastic sliding law,
\begin{equation}
    \label{eqn:pseudoplastic}
    \vec{\tau}_b = -\tau_c \frac{\vec{v}_b}{{v_{th}}^q\,|\vec{v}_b|^{1-q}} \,,
\end{equation}
relates the bed-parallel shear stresses, $\vec{\tau}_b$, to the sliding
velocity, $\vec{v}_b$ (Table~\ref{tab:params}).
The yield stress, $\tau_c$,
is modelled using the Mohr--Coulomb criterion,
\begin{equation}
   \tau_c = c_0 + N\,\tan{\phi} \,,
\end{equation}
where cohesion, $c_0$, is assumed to be zero. The friction angle, $\phi$,
varies from 15 to 45\degree as a piecewise-linear function of modern bed
elevation, with the lowest value occuring below modern sea level (0\,m above
sea level, m~a.s.l.) and the highest value occuring above the generalised
elevation of the highest shorelines
\citep[200\,m~a.s.l.,][Fig.~5]{Clague.1981}, thus accounting for
a~weakening of till associated with the presence of marine sediments
(cf. \citealp{Martin.etal.2011}; \citealp[supplement]{Aschwanden.etal.2013};
\citealp{PISM-authors.2014}). Effective pressure, $N$,
is related to the ice overburden stress, $\rho gh$, and the modelled amount of
subglacial water, using a formula derived from laboratory experiments with till
extracted from an Antarctic ice stream \citep{Tulaczyk.etal.2000,
Bueler.Pelt.2014},
\begin{equation}
    N = \delta \rho gh \, 10^{(e_0/C_c) (1 - (W/W_{max}))} \,,
\end{equation}
where $\delta$ is chosen as 0.02, $e_0$ is a measured reference void ratio and
$C_c$ a measured compressibility coefficient (Table~\ref{tab:params}). The
amount of water at the base, $W$, varies from zero to $W_{max}=2$\,m, a
threshold above which instantaneous drainage is assumed.
Finally, the bedrock topography responds to ice load
following a bedrock deformation model that includes local isostasy,
plastic lithosphere flexure and viscous astenosphere deformation in an infinite
half-space
\citep[Table~\ref{tab:params};][]{Lingle.Clark.1985,Bueler.etal.2007}.
A relatively low viscosity value of $\nu_m = 1\times10^{19}$\,\unit{Pa\,s} is
used for the astenosphere in accordance to the results of regional glacial
isostatic adjustment modelling at the northern Cascadia subduction zone
\citep{James.etal.2009}.

@@

\begin{figure}
  \includegraphics{atm}
  \caption{Monthly mean near-surface air temperature, precipitation, and
           standard deviation of daily mean temperature for January and July
           from the North American Regional Reanalysis
           \citep[NARR;][]{Mesinger.etal.2006},
           used to force the ice sheet model. Note the
           strong contrasts in seasonality, timing of the precipitation peak,
           and temperature variability over the model domain, notably between
           coastal and inland regions.}
  \label{fig:atm}
\end{figure}

Ice surface accumulation and ablation are computed from monthly mean
near-surface air temperature, $T_m$, monthly standard deviation of near-surface
air temperature, $\sigma$, and monthly precipitation, $P_m$, using
a~temperature-index model \citep[e.g.,][]{Hock.2003}.
Accumulation is equal to
precipitation when air temperatures are below 0\,\unit{{\degree}C}, and
decreases to zero linearly with temperatures between 0 and
2\,\unit{{\degree}C}. Ablation is computed from the number of positive
degree-days (PDD), defined as the integral of temperatures above
0\,\unit{{\degree}C} in one year.

@@

While simulations driven by the GRIP and the two Antarctic palaeo-temperature
records attain a last ice volume maximum between 19.1 and 16.8\,ka, those
informed by the NGRIP and the two oceanic palaeo-temperature records attain their
maximum ice volumes thousands of years earlier. Moreover, the ODP~1012 run
yields a rapid deglaciation of the modelled area prior to 17\,ka. The ODP~1020
simulation predicts an early maximum in ice volume at 23.2\,ka, followed by
slower deglaciation than modelled using the other palaeo-temperature records.
Finally, model runs forced by Antarctic palaeo-temperature records result in a
rapid and uninterrupted deglaciation after the last glacial maximum, whereas the
simulation driven by the GRIP palaeo-temperature record also results in a rapid
deglaciation but in three steps, separated by two ice volume standstills
(Fig.~\ref{fig:timeseries}).

@@

Among the simulations presented here, only those forced with the GRIP, EPICA
and Vostok palaeo-temperature records yield Cordilleran ice sheet maximum
extents that may be compatible with these field constraints
(Fig.~\ref{fig:timeseries}, lower panel; Table~\ref{tab:extrema}).
Simulations driven by the NGRIP, ODP~1012 and ODP~1020
palaeo-temperature records, on the contrary, yield MIS~2 maximum Cordilleran
ice sheet extents that pre-date field-based constraints by several thousands of
years. Concerning the simulations driven by oceanic records, this early
deglaciation is caused by an early warming present in the alkenone
palaeo-temperature reconstructions (Fig.~\ref{fig:timeseries}, upper panel;
\citealp[Fig.~3]{Herbert.etal.2001}). However, this
early warming is a local effect, corresponding to a weakening of the California
current \citep{Herbert.etal.2001}. The California current, driving cold
waters southwards along the south-western coast of North America,
has been shown to have weakened during each peak global glaciation (in SPECMAP)
during the past 550\,ka, including the LGM, resulting in paradoxically warmer
sea-surface temperatures at the locations of the ODP~1012 and ODP~1020 sites
\citep{Herbert.etal.2001}.

@@

Similarly, the simulation using the NGRIP palaeo-temperature record depicts an
early onset of deglaciation (Fig.~\ref{fig:timeseries}) following its last
glacial volume maximum attained about 6\,ka earlier (22.9\,ka,
Table~\ref{tab:extrema}) than dated evidence of the last glacial maximum
advance. Hence, we focus the rest of our analysis on simulations forced by
palaeo-temperature records from the GRIP and EPICA ice cores that appear to
produce the most realistic reconstructions of regional glaciation history among
all records tested. To allow for a more
detailed comparison against the geologic record, these two simulations were
re-run using the high-resolution grid (Fig.~\ref{fig:timeseries}, lower panel,
dotted lines).

@@

This could reflect the fact that our model does not include
feedback mechanisms between ice sheet topography and the regional climate.
Firstly, during the build-up phase preceding the LGM, rapid accumulation over
the Coast Mountains enhanced the topographic barrier formed by these mountain
ranges, which likely resulted in a decrease of precipitation and, therefore,
accumulation in the interior. Secondly, latent warming of the moisture-depleted
air parcels flowing over this enhanced topography could have resulted in an
inflow of potentially warmer air over the eastern flank of the ice sheet,
increasing melt
along the advancing margin \citep[cf.][]{Langen.etal.2012}. Because these two
processes, both with a tendency to limit ice-sheet growth, are absent from
our model, the
eastern margin of the ice sheet and the position of the main meridional ice
divide are certainly biased towards the east in our simulations
\citep{Seguinot.etal.2014}.

@@

The modelled ice sheet configurations corresponding to ice volume maxima during
MIS~4 are more sensitive to the choice of atmospheric forcing than those
corresponding to ice volume maxima during MIS~2 (Fig.~\ref{fig:snapshots},
upper panels;
Fig.~\ref{fig:mis4}). The GRIP simulation (Fig.~\ref{fig:mis4}, left panel)
results in a modelled maximum ice sheet extent that closely resembles that
obtained
during MIS~2, with the only major difference being slightly less extensive
northern and eastern sectors. In contrast, the EPICA simulation produces a lower
ice volume maximum (Fig.~\ref{fig:timeseries}), which translates in the
modelled ice sheet geometry into a significantly reduced southern sector, more
restricted ice cover in northern and eastern sectors, and generally lower
ice surface elevations in the interior (Fig.~\ref{fig:mis4}, left panel).
Thus, only the GRIP simulation can explain the presence of ice-rafted debris
in the marine sediment record offshore Vancouver Island at
ca.\,\unit{\chem{^{14}C}\,cal\,ka} \citep{Cosma.etal.2008}.

@@

\begin{figure*}
  \includegraphics{duration}
  \caption{Modelled duration of ice cover during the last 120\,ka.
           Note the irregular colour scale. A continuous ice cover spanning
           from the Alaska Range (AR) to the southern Coast Mountains (CM) and
           the Columbia and Rocky mountains (RM) exists for about 32\,ka in
           the GRIP simulation and 26\,ka in the EPICA simulation.
           The maximal extent of the ice sheet generally
           corresponds to relatively short durations of ice cover, but ice
           cover persists over the Skeena Mountains (SM) during most of the
           simulation.}
  \label{fig:duration}
\end{figure*}

Palaeo-glaciological reconstructions are generally more robust
for maximum ice sheet extents and late ice sheet configurations than for
intermediate or minimum ice sheet extents and older ice sheet configurations
\citep{Kleman.etal.2010}. However,
these maximum stages are, by nature, extreme configurations, which do not
necessarily represent the dominant patterns of glaciation throughout the period
of ice cover \citep{Porter.1989, Kleman.Stroeven.1997}.

@@

It is generally believed that the Cordilleran ice sheet formed by the
coalescence of several mountain-centred ice caps \citep{Davis.Mathews.1944}.
In our simulations, major ice-dispersal centres, visible on the modelled ice
cover duration maps (Fig.~\ref{fig:duration}), are located over the Coast
Mountains (CM), the Columbia and Rocky mountains (CRM), the Skeena Mountains
(SM), and the Selwyn and Mackenzie Mountains (SMKM). Although the Coast,
Columbia, Rocky and Skeena mountains (CM, CRM, SM) are covered by mountain
glaciers for most of the last glacial cycle, providing durable nucleation
centres for an ice sheet, this is not the case for the Selwyn and
Mackenzie mountains (SMKM), where ice cover on the highest peaks is limited to a
fraction of the last glacial cycle. In other words, the Selwyn and Mackenzie
mountains appear as a secondary ice-dispersal centre which ``activates'' only
during the coldest periods of the last glacial cycle. The Northern Rocky
Mountains (Fig.~\ref{fig:duration}, NRM), do not act as a nucleation centre,
but rather as a pinning point for the Cordilleran ice sheet margin coming from
the west.

Perhaps the most striking feature displayed by the distributions of modelled
ice cover is the persistence of the Skeena Mountains ice cap throughout the
entire last glacial cycle and its predominance over the other ice-dispersal
centres
(Figs.~\ref{fig:snapshots} and~\ref{fig:duration}). Regardless of the applied
forcing, this ice cap appears to survive MIS~3 (Fig.~\ref{fig:snapshots},
middle panels), and serves as a nucleation centre at the onset of the glacial
readvance towards the LGM (MIS~2). For the neighbouring Laurentide ice sheet,
the importance of residual ice for North
American glacial history leading up to the LGM has been illustrated by the
MIS~3 residual ice bodies in northern and eastern Canada as nucleation centres
for a much more extensive MIS~2 configuration \citep{Kleman.etal.2010}.

@@

\begin{figure*}
  \vspace{-1.5mm}  % enough to allow the two figures to fit on one page
  \includegraphics{deglac}
  \caption{Modelled age of the last deglaciation. Areas where the MIS~4 glacial
           advance exceeded the last glacial maximum advanced are marked in
           green. Hatches denote re-advance of mountain-centred ice caps and
           the decaying ice sheet between 14 and 10\,ka., which is more
           pronounced in the GRIP-driven simulation.
           Dashed segments indicate the location of profiles used in
           Figs.~\ref{fig:profiles-grip} and~\ref{fig:profiles-epica}.}
  \label{fig:deglac}
\end{figure*}

@@

\begin{figure*}
  \includegraphics{lastflow}
  \caption{Modelled of the deglacial basal ice velocities. Hatches
           indicate areas that remain non-sliding throughout deglaciation
           (22.0--8.0\,ka), notably including the Interior Plateau (IP).
           Note the concentric patterns of deglacial flow in the Liard
           Lowland (LL).
           Sliding grid cells were distinguished from non-sliding grid cells
           using a velocity threshold of 1\,\unit{m\,yr^{-1}}.
           Dashed segments indicate the location of profiles used in
           Figs.~\ref{fig:profiles-grip} and~\ref{fig:profiles-epica}.}
  \label{fig:lastflow}
\end{figure*}

>>>>

More than a century of exploration and geological investigation of the
Cordilleran mountains have led to many observations in support of the former
ice sheet
    \citep{Jackson.Clague.1991}.
Despite the lack of documented end moraines offshore, in the zone of confluence
with the Laurentide ice sheet, and in areas swept by the Missoula floods
    \citep{Carrara.etal.1996},
moraines that demarcate the northern and south-western margins provide key
constraints that allow reasonable reconstructions of maximum ice sheet extents
    (\citealp{Prest.etal.1968}; \citealp[Fig. 1.12]{Clague.1989};
     \citealp{Duk-Rodkin.1999};
     \citealp{Booth.etal.2003}; \citealp{Dyke.2004}).
The LGM Cordilleran ice sheet extent was short-lived, as indicated by
field evidence from radiocarbon dating
    \citep{Clague.etal.1980, Clague.1985, Clague.1986, Porter.Swanson.1998,
           Menounos.etal.2008},
cosmogenic exposure dating
    \citep{Stroeven.etal.2010, Stroeven.etal.2014, Margold.etal.2014},
bedrock deformation in response to former ice loads
    \citep{Clague.James.2002, Clague.etal.2005},
and offshore sedimentary records
    \citep{Cosma.etal.2008, Davies.etal.2011}.
However, former ice thicknesses and, therefore, the ice sheet's contribution to
the LGM sea level lowstand
    \citep{Carlson.Clark.2012, Clark.Mix.2002}
remain uncertain.

@@

\begin{figure}
  \includegraphics{locmap}
  \caption{Relief map of northern North America showing a reconstruction of the
           areas once covered by the Cordilleran (CIS), Laurentide (LIS),
           Innuitian (IIS) and Greenland (GIS) ice sheets during the last
           18\,\unit{\chem{^{14}C}\,ka} \citep[21.4\,cal\,ka,][]{Dyke.2004}.
           The rectangular box denotes the location of the
           modelling domain used in this study. Major mountain ranges covered
           by the ice sheet include the Alaska Range (AR), the Wrangell and
           St.-Elias mountains (WSEM), the Selwyn and Mackenzie mountains
           (SMKM), the Skeena Mountains (SM), the Coast and Rocky
           Mountains (CRM) and the North Cascades (NC). Major depressions
           include the Liard Lowland (LL), the Interior Plateau of British
           Columbia (IP) and the Puget Lowland (PL). The background
           map consists of ETOPO1 \citep{Amante.Eakins.2009} and Natural Earth
           Data \citep{Patterson.Kelso.2015}.}
  \label{fig:locmap}
\end{figure}

In contrast, evidence for the deglaciation history of the Cordilleran
ice sheet since the LGM is considerable, albeit mostly at a regional scale.
Geomorphological evidence from south-central British Columbia indicates a rapid
deglaciation, including an early emergence of the elevated areas while thin,
stagnant ice still covered the surrounding lowlands
    \citep{Fulton.1967, Fulton.1991, Margold.etal.2011, Margold.etal.2013a}.
This model, although credible, may not apply in all areas of the Cordilleran
ice sheet
    \citep{Margold.etal.2013}.
Although solid evidence for late-glacial glacier re-advances have been found in
the Coast, Columbia and Rocky mountains
    \citep{Clague.etal.1997, Friele.Clague.2002, Friele.Clague.2002a,
           Kovanen.2002, Kovanen.Easterbrook.2002, Lakeman.etal.2008,
           Menounos.etal.2008},
it appears to be sparser than for formerly glaciated regions surrounding
the North Atlantic
    \citep[e.g.,][]{Sissons.1979, Lundqvist.1987,
                    Ivy-Ochs.etal.1999, Stea.etal.2011}.
Nevertheless, recent oxygen isotope measurements from Gulf of Alaska sediments
reveals a climatic evolution highly correlated to that of Greenland during this
period, including a distinct late glacial cold reversal between 14.1 and
11.7\,ka \citep{Praetorius.Mix.2014}.

In general, the topographic complexity of the North American Cordillera and its
effect on glacial history have inhibited the reconstruction of ice sheet-wide
glacial advance and retreat patterns such as
those available for the Fennoscandian and Laurentide ice sheets
     \citep{Boulton.etal.2001, Dyke.Prest.1987, Dyke.etal.2003,
            Kleman.etal.1997, Kleman.etal.2010, Stroeven.etal.inreview}.
Here, we use a numerical ice sheet model \citep{PISM-authors.2015},
calibrated against field-based evidence, to perform a quantitative
reconstruction of the Cordilleran ice sheet history through the last glacial
cycle, and
analyse some of the long-standing questions related to its evolution:

@@

Although numerical ice sheet modelling has been established as a useful tool to
improve our understanding of the Cordilleran ice sheet
    (\citealp[p.~227]{Jackson.Clague.1991}; \citealp{Robert.1991},
     \citealp{Marshall.etal.2000}),
the ubiquitously mountainous
topography of the region has presented two major challenges to its application.
First, only recent development of numerical ice sheet models and underlying
scientific computing tools \citep{Bueler.Brown.2009, Balay.etal.2015} has
allowed for high-resolution numerical modelling of glaciers and ice sheets on
mountainous terrain
over millennial time scales \citep[e.g.,][]{Golledge.etal.2012}. Second, the
complex
topography of the North American Cordillera also induces strong gradients in
temperature
seasonality and distinct patterns of precipition, thus requiring the use of
high-resolution climate forcing fields as an input to the ice sheet model
\citep{Seguinot.etal.2014}.

@@

The simulations presented here were run using the Parallel Ice Sheet Model
(PISM, development version~8ff7cbe), an open source,
finite difference, shallow ice sheet model \citep{PISM-authors.2015}. The model
requires input on basal topography, sea level, geothermal heat flux and
climate forcing. It computes the evolution of ice extent
and thickness over time, the thermal and dynamic
states of the ice sheet, and the associated lithospheric response.

@@

A~pseudo-plastic sliding law,
\begin{equation}
    \label{eqn:pseudoplastic}
    \vec{\tau}_b = -\tau_c \frac{\vec{v}_b}{{v_{th}}^q\,|\vec{v}_b|^{1-q}} \,,
\end{equation}
relates the bed-parallel shear stresses, $\vec{\tau}_b$, to the sliding
velocity, $\vec{v}_b$ (Table~\ref{tab:params}).
The yield stress, $\tau_c$,
is modelled using the Mohr--Coulomb criterion,
\begin{equation}
   \tau_c = c_0 + N\,\tan{\phi} \,,
\end{equation}
where cohesion, $c_0$, is assumed to be zero. The friction angle, $\phi$,
varies from 15 to 45\degree as a piecewise-linear function of modern bed
elevation, with the lowest value occuring below modern sea level (0\,m above
sea level, m~a.s.l.) and the highest value occuring above the generalised
elevation of the highest shorelines
\citep[200\,m~a.s.l.,][Fig.~5]{Clague.1981}, thus accounting for
a~weakening of till associated with the presence of marine sediments
(cf. \citealp{Martin.etal.2011}; \citealp[supplement]{Aschwanden.etal.2013};
\citealp{PISM-authors.2015}). Effective pressure, $N$,
is related to the ice overburden stress, $\rho gh$, and the modelled amount of
subglacial water, using a formula derived from laboratory experiments with till
extracted from an Antarctic ice stream \citep{Tulaczyk.etal.2000,
Bueler.Pelt.2014},
\begin{equation}
    N = \delta \rho gh \, 10^{(e_0/C_c) (1 - (W/W_{max}))} \,,
\end{equation}
where $\delta$ is chosen as 0.02, $e_0$ is a measured reference void ratio and
$C_c$ a measured compressibility coefficient (Table~\ref{tab:params}). The
amount of water at the base, $W$, varies from zero to $W_{max}=2$\,m, a
threshold above which any additional melt water is assumed to drain off
instantaneously. Finally, the bedrock topography responds to ice load
following a bedrock deformation model that includes local isostasy,
elastic lithosphere flexure and viscous astenosphere deformation in an infinite
half-space
\citep[Table~\ref{tab:params};][]{Lingle.Clark.1985,Bueler.etal.2007}.
A relatively low viscosity value of $\nu_m = 1\times10^{19}$\,\unit{Pa\,s} is
used for the astenosphere in accordance to the results of regional glacial
isostatic adjustment modelling at the northern Cascadia subduction zone
\citep{James.etal.2009}.

@@

\begin{figure}
  \includegraphics{atm}
  \caption{Monthly mean near-surface air temperature, precipitation, and
           standard deviation of daily mean temperature for January and July
           from the North American Regional Reanalysis
           \citep[NARR;][]{Mesinger.etal.2006}, used to force the surface mass
           balance (PDD) component of the ice sheet model. Note the
           strong contrasts in seasonality, timing of the precipitation peak,
           and temperature variability over the model domain, notably between
           coastal and inland regions.}
  \label{fig:atm}
\end{figure}

Ice surface accumulation and ablation are computed from monthly mean
near-surface air temperature, $T_m$, monthly standard deviation of near-surface
air temperature, $\sigma$, and monthly precipitation, $P_m$, using
a~temperature-index model \citep[e.g.,][]{Hock.2003}.
Accumulation is equal to
precipitation when air temperatures are below 0\,\unit{{\degree}C}, and
decreases to zero linearly with temperatures between 0 and
2\,\unit{{\degree}C}. Ablation is computed from the PDD, defined as the
integral of temperatures above 0\,\unit{{\degree}C} in one year.

@@

While simulations driven by the GRIP and the two Antarctic palaeo-temperature
records attain a last ice volume maximum between 19.1 and 16.8\,ka, those
informed by the NGRIP and the two oceanic palaeo-temperature records attain their
maximum ice volumes thousands of years earlier. Moreover, the ODP~1012 run
yields a rapid deglaciation of the modelled area prior to 17\,ka. The ODP~1020
simulation predicts an early maximum in ice volume at 23.2\,ka, followed by
slower deglaciation than modelled using the other palaeo-temperature records.
Finally, model runs forced by Antarctic palaeo-temperature records result in a
rapid and uninterrupted deglaciation after the last glacial maximum, whereas the
simulation driven by the GRIP palaeo-temperature record also results in a rapid
deglaciation but in three steps, separated by two periods of ice volume increase
(Fig.~\ref{fig:timeseries}).

@@

Among the simulations presented here, only those forced with the GRIP, EPICA
and Vostok palaeo-temperature records yield Cordilleran ice sheet maximum
extents that may be compatible with these field constraints
(Fig.~\ref{fig:timeseries}, lower panel; Table~\ref{tab:extrema}).
Simulations driven by the NGRIP, ODP~1012 and ODP~1020
palaeo-temperature records, on the contrary, yield MIS~2 maximum Cordilleran
ice sheet extents that pre-date field-based constraints by several thousands of
years (about 6, 6 and 4\,ka respectively).
Concerning the simulations driven by oceanic records, this early
deglaciation is caused by an early warming present in the alkenone
palaeo-temperature reconstructions (Fig.~\ref{fig:timeseries}, upper panel;
\citealp[Fig.~3]{Herbert.etal.2001}). However, this
early warming is a local effect, corresponding to a weakening of the California
current \citep{Herbert.etal.2001}. The California current, driving cold
waters southwards along the south-western coast of North America,
has been shown to have weakened during each peak global glaciation (in SPECMAP)
during the past 550\,ka, including the LGM, resulting in paradoxically warmer
sea-surface temperatures at the locations of the ODP~1012 and ODP~1020 sites
\citep{Herbert.etal.2001}.

@@

Similarly, the simulation using the NGRIP palaeo-temperature record depicts an
early onset of deglaciation (Fig.~\ref{fig:timeseries}) following its last
glacial volume maximum attained about 6\,ka earlier (22.9\,ka,
Table~\ref{tab:extrema}) than dated evidence of the last glacial maximum
advance. In addition, there exist a fair agreement between the EPICA and Vostok
alaeo-temperature records, resulting in only little differences between the
simulations driven by those records. These differences are not analysed
further; instead we focus the rest of our analysis on simulations forced by
palaeo-temperature records from the GRIP and EPICA ice cores that appear to
produce the most realistic reconstructions of regional glaciation history among
all records tested, yet bearing significant disparities. To allow for a more
detailed comparison against the geologic record, these two simulations were
re-run using the high-resolution grid (Fig.~\ref{fig:timeseries}, lower panel,
dotted lines).

@@

This could reflect the fact that our model does not include
feedback mechanisms between ice sheet topography and the regional climate.
Firstly, during the build-up phase preceding the LGM, rapid accumulation over
the Coast Mountains enhanced the topographic barrier formed by these mountain
ranges, which likely resulted in a decrease of precipitation and, therefore,
a decrease of
accumulation in the interior. Secondly, latent warming of the moisture-depleted
air parcels flowing over this enhanced topography could have resulted in an
inflow of potentially warmer air over the eastern flank of the ice sheet,
increasing melt
along the advancing margin \citep[cf.][]{Langen.etal.2012}. Because these two
processes, both with a tendency to limit ice-sheet growth, are absent from
our model, the
eastern margin of the ice sheet and the position of the main meridional ice
divide are certainly biased towards the east in our simulations
\citep{Seguinot.etal.2014}.

@@

The modelled ice sheet configurations corresponding to ice volume maxima during
MIS~4 are more sensitive to the choice of atmospheric forcing than those
corresponding to ice volume maxima during MIS~2 (Fig.~\ref{fig:snapshots},
upper panels;
Fig.~\ref{fig:mis4}). The GRIP simulation (Fig.~\ref{fig:mis4}, left panel)
results in a modelled maximum ice sheet extent that closely resembles that
obtained
during MIS~2, with the only major difference being slightly less extensive
northern and eastern sectors. In contrast, the EPICA simulation produces a lower
ice volume maximum (Fig.~\ref{fig:timeseries}), which translates in the
modelled ice sheet geometry into a significantly reduced southern sector, more
restricted ice cover in northern and eastern sectors, and generally lower
ice surface elevations in the interior (Fig.~\ref{fig:mis4}, left panel).
Thus, only the GRIP simulation can explain the presence of ice-rafted debris
in the marine sediment record offshore Vancouver Island at
ca.~45\,\unit{\chem{^{14}C}\,cal\,ka} \citep{Cosma.etal.2008}.

@@

\begin{figure*}
  \includegraphics{duration}
  \caption{Modelled duration of ice cover during the last 120\,ka.
           Note the irregular colour scale. A continuous ice cover spanning
           from the Alaska Range (AR) to the southern Coast Mountains (CM) and
           the Columbia and Rocky mountains (RM) exists for about 32\,ka in
           the GRIP simulation and 26\,ka in the EPICA simulation.
           The maximum extent of the ice sheet generally
           corresponds to relatively short durations of ice cover, but ice
           cover persists over the Skeena Mountains (SM) during most of the
           simulation. See Fig.~\ref{locmap} for a list of abbreviations.}
  \label{fig:duration}
\end{figure*}

Palaeo-glaciological reconstructions are generally more robust
for maximum ice sheet extents and late ice sheet configurations than for
intermediate or minimum ice sheet extents and older ice sheet configurations
\citep{Kleman.etal.2010}. However,
these maximum stages are, by nature, extreme configurations, which do not
necessarily represent the dominant patterns of glaciation throughout the period
of ice cover \citep{Porter.1989, Kleman.Stroeven.1997, Kleman.etal.2008,
Kleman.etal.2010}.

@@

It is generally believed that the Cordilleran ice sheet formed by the
coalescence of several mountain-centred ice caps \citep{Davis.Mathews.1944}.
In our simulations, major ice-dispersal centres, visible on the modelled ice
cover duration maps (Fig.~\ref{fig:duration}), are located over the Coast
Mountains (CM), the Columbia and Rocky mountains (CRM), the Skeena Mountains
(SM), and the Selwyn and Mackenzie mountains (SMKM). Although the Coast,
Columbia, Rocky and Skeena mountains (CM, CRM, SM) are covered by mountain
glaciers for most of the last glacial cycle, providing durable nucleation
centres for an ice sheet, this is not the case for the Selwyn and
Mackenzie mountains (SMKM), where ice cover on the highest peaks is limited to a
fraction of the last glacial cycle. In other words, the Selwyn and Mackenzie
mountains appear as a secondary ice-dispersal centre which is only active
during the coldest periods of the last glacial cycle. The Northern Rocky
Mountains (Fig.~\ref{fig:duration}, NRM) do not act as a nucleation centre,
but rather as a pinning point for the Cordilleran ice sheet margin coming from
the west.

Perhaps the most striking feature displayed by the distributions of modelled
ice cover is the persistence of the Skeena Mountains ice cap throughout the
entire last glacial cycle and its predominance over the other ice-dispersal
centres
(Figs.~\ref{fig:snapshots} and~\ref{fig:duration}, SM). Regardless of the applied
forcing, this ice cap appears to survive MIS~3 (Fig.~\ref{fig:snapshots},
middle panels), and serves as a nucleation centre at the onset of the glacial
readvance towards the LGM (MIS~2). This situation appear similar to the neighbouring Laurentide ice sheet, for which the importance of residual ice for
glacial history leading up to the LGM has been illustrated by the
MIS~3 residual ice bodies in northern and eastern Canada as nucleation centres
for a much more extensive MIS~2 configuration \citep{Kleman.etal.2010}.

@@

\begin{figure*}
  \vspace{-1.5mm}  % enough to allow the two figures to fit on one page
  \includegraphics{deglac}
  \caption{Modelled age of the last deglaciation. Areas where the MIS~4 glacial
           advance exceeded the last glacial maximum advanced are marked in
           green. Hatches denote re-advance of mountain-centred ice caps and
           the decaying ice sheet between 14 and 10\,ka., which is very
           restricted in the EPICA-driven simulation.
           Dashed segments indicate the location of profiles used in
           Figs.~\ref{fig:profiles-grip} and~\ref{fig:profiles-epica}.}
  \label{fig:deglac}
\end{figure*}

@@

\begin{figure*}
  \includegraphics{lastflow}
  \caption{Modelled deglacial basal ice velocities. Hatches
           indicate areas that remain non-sliding throughout deglaciation
           (22.0--8.0\,ka), notably including parts of the Interior Plateau (IP).
           Note the concentric patterns of deglacial flow in the Liard
           Lowland (LL).
           Sliding grid cells were distinguished from non-sliding grid cells
           using a velocity threshold of 1\,\unit{m\,yr^{-1}}.
           Dashed segments indicate the location of profiles used in
           Figs.~\ref{fig:profiles-grip} and~\ref{fig:profiles-epica}.}
  \label{fig:lastflow}
\end{figure*}
